\pkg{dataMaid}: your assistant for data cleaning in \proglang{R}

### 2017-01-30 latex/article_vol2.Rnw
Fix package reference
DataExplorer
Rewritten part 1

\section{Introduction}
Though data cleaning might be regarded as a somewhat tedious activity,
adequate data cleaning is crucial in any data analysis. With
ever-growing dataset sizes and complexities, statisticians and data
analysts find themselves spending a large portion of their time on
data cleaning and on data wrangling. While a computer should never
make unsupervised decisions on what should be done to potential
errors in the dataset, it can still be an extremely useful tool in the
data cleaning process. Some errors can be tracked down and flagged by a
computer without further ado, while other types of errors need an analytic
context in order to shine through. But even in the latter case, well-designed
software can aid the process tremendously by giving the human investigator the
information needed for identifying issues.

Online tools such as OpenRefine (\url{http://openrefine.org/}) and
\proglang{R}-packages such as \pkg{plyr} \citep{plyr}, and
\pkg{data.table} \citep{data.table} have made data wrangling a lot
easier, but only a handful of packages such as \pkg{editrules}
\citep{editrules}, \pkg{validate} \citep{validate}, \pkg{DataCombine}
\citep{DataCombine}, \pkg{janitor} \citep{janitor}, and
\pkg{DescTools} \citep{DescTools} attempt to implement systematic,
reproducible data cleaning.  These packages use different approaches
for data cleaning: \pkg{editrules} and \pkg{validate} provide
frameworks for setting up and checking constraints on the variables,
while \pkg{DataCombine} and \pkg{janitor} both provide a few functions
for identifying problems (e.g, duplicates, dates coded as numbers,
etc.) in data.

@@

\subsubsection{isID --- a new checkFunction without problem values}
First, let's define the \code{isID()} function. As this function is not
supposed to list problematic values, it falls within
the category of \code{checkFunctions} represented by
\code{myFullVarCheckFunction()} in the above. We do not particularly wish
to use this function interactively, so we will stick to the minimal
requirements of a \code{checkFunction} used in \code{check()} (see
Table~\ref{table.functionTypes}). The function can then be defined by

@@

As \code{mosaicVisual} is now a full-blooded \code{visualFunction}, it
will also be included in the \code{allVisualFunctions()} output table:
<<>>=
allVisualFunctions()
@

Now that we are done with the definition of \code{mosaicVisual()} we can
turn to the next function in line, \code{countZeros}.

@@

As with the previous two functions, we also change its class. Note,
however, that for \code{checkFunction}s, the function description will
appear in the document produced by \code{clean()} (in the \textit{Data
  cleaning summary} section of the output), so now this is not only
done for the sake of the \code{allCheckFunctions()} output.

\subsubsection{Calling the new summarize/visualize/check functions from clean()}
Now, we are ready to use these new functions in a \code{clean()}
call. The extended \pkg{dataMaid} output document should have the
following modifications, relative to the standard \pkg{dataMaid} output:
\begin{itemize}
\item We want to add the new pre-check function, \code{isID}, to the already existing pre-checks.
\item We wish to change the plot type for all variables to the new mosaic plot.
\item We want the new summary function, \code{countZeros}, to be added to the summaries performed on all variable types but \code{Date} and \code{logical}.
\item We want the new check function, \code{identifyColon}, to be added to the checks performed on \code{character}, \code{factor} and \code{labelled} variables.
\end{itemize}
These options are specified as follows: %\hl{Jeg synes ikke testData
                                %egner sig her. Der er ingen koloner
                                %og CPRvar er et mærkeligt, internt
                                %navn, hvis ikke man kender danske
                                %CPR-numre. Langt de fleste variable
                                %har desuden ingen nuller. Måske skal
                                %der laves et nyt datasæt?}
<<eval=FALSE>>=
data("exampleData")
clean(exampleData,
  preChecks = c("isKey", "isEmpty", "isID"),
  allVisuals = "mosaicVisual",
  characterSummaries = c(defaultCharacterSummaries(), "countZeros"),
  factorSummaries = c(defaultFactorSummaries(), "countZeros"),
  labelledSummaries = c(defaultLabelledSummaries(), "countZeros"),
  numericSummaries = c(defaultNumericSummaries(), "countZeros"),
  integerSummaries = c(defaultIntegerSummaries(), "countZeros"),
  characterChecks = c(defaultCharacterChecks(), "identifyColons"),
  factorChecks = c(defaultFactorChecks(), "identifyColons"),
  labelledCheck = c(defaultLabelledChecks(), "identifyColons"))
@

>>>>

\section{Introduction}
Though data cleaning might be regarded as a somewhat tedious activity,
adequate data cleaning is crucial in any data analysis. With
ever-growing dataset sizes and complexities, statisticians and data
analysts find themselves spending a large portion of their time on
data cleaning and on data wrangling. While a computer should never
make unsupervised decisions on what should be done to potential
errors in a dataset, it can still be an extremely useful tool in the
data cleaning process. Some errors can be tracked down and flagged by a
computer without further ado, while other types of errors need an analytic
context in order to shine through. But even in the latter case, well-designed
software can aid the process tremendously by giving the human investigator the
information needed for identifying issues.

Online tools such as OpenRefine (\url{http://openrefine.org/}) and
\proglang{R}-packages such as \pkg{plyr} \citep{plyr}, and
\pkg{data.table} \citep{data.table} have made data wrangling a lot
easier, but only a handful of packages such as \pkg{editrules}
\citep{editrules}, \pkg{validate} \citep{validate}, \pkg{DataCombine}
\citep{DataCombine}, \pkg{janitor} \citep{janitor}, and
\pkg{DataExplorer} \citep{DataExplorer} attempt to implement systematic,
reproducible data cleaning.  These packages use different approaches
for data cleaning: \pkg{editrules} and \pkg{validate} provide
frameworks for setting up and checking constraints on the variables,
while \pkg{DataCombine} and \pkg{janitor} both provide a few functions
for identifying problems (e.g, duplicates, dates coded as numbers,
etc.) in data.

@@

\subsubsection{isID --- a new checkFunction without problem values}
First, let's define the \code{isID()} function. As this function is not
supposed to list problematic values, it falls within
the category of \code{checkFunction}s represented by
\code{myFullVarCheckFunction()} in the above. We do not particularly wish
to use this function interactively, so we will stick to the minimal
requirements of a \code{checkFunction} used in \code{check()} (see
Table~\ref{table.functionTypes}). The function can then be defined by

@@

As \code{mosaicVisual()} is now a full-blooded \code{visualFunction}, it
will also be included in the \code{allVisualFunctions()} output table:
<<>>=
allVisualFunctions()
@

Now that we are done with the definition of \code{mosaicVisual()} we can
turn to the next function in line, \code{countZeros()}.

@@

Note, that for \code{checkFunction}s, the function description will
appear in the document produced by \code{clean()} (in the \textit{Data
  cleaning summary} section of the output), so for this type of
function, the change of class is not only done for the sake of the
\code{allCheckFunctions()} output.

\subsubsection{Calling the new summarize/visualize/check functions from clean()}
Now, we are ready to use these new functions in a \code{clean()}
call. The extended \pkg{dataMaid} output document should have the
following modifications, relative to the standard \pkg{dataMaid} output:
\begin{itemize}
\item We want to add the new pre-check function, \code{isID()}, to the already existing pre-checks.
\item We wish to change the plot type for all variables to the new mosaic plot.
\item We want the new summary function, \code{countZeros()}, to be added to the summaries performed on all variable types but \code{Date} and \code{logical}.
\item We want the new check function, \code{identifyColons()}, to be added to the checks performed on \code{character}, \code{factor} and \code{labelled} variables.
\end{itemize}
These options are specified as follows:

<<eval=FALSE>>=
data("exampleData")
clean(exampleData,
  preChecks = c("isKey", "isEmpty", "isID"),
  allVisuals = "mosaicVisual",
  characterSummaries = c(defaultCharacterSummaries(), "countZeros"),
  factorSummaries = c(defaultFactorSummaries(), "countZeros"),
  labelledSummaries = c(defaultLabelledSummaries(), "countZeros"),
  numericSummaries = c(defaultNumericSummaries(), "countZeros"),
  integerSummaries = c(defaultIntegerSummaries(), "countZeros"),
  characterChecks = c(defaultCharacterChecks(), "identifyColons"),
  factorChecks = c(defaultFactorChecks(), "identifyColons"),
  labelledCheck = c(defaultLabelledChecks(), "identifyColons"))
@
